Maximum Entropy Prediction Markets

### 2014-02-08 WriteUp/maxentpredmkts.tex
completed first draft of exp utility section

%new commands
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\mX}{\mathcal{X}}
\newcommand{\bR}{\mathbf{R}}
\newcommand{\grad}{\nabla}
\newcommand{\Exp}{\mathbf{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\mM}{\ensuremath{\mathcal{M}}}
\newcommand{\hmu}{\ensuremath{\hat{\mu}}}
\newcommand{\mP}{\ensuremath{\mathcal{P}}}
\newcommand{\mY}{\ensuremath{\mathcal{Y}}}
\newcommand{\mS}{\ensuremath{\mathcal{S}}}
\newcommand{\ds}{\displaystyle}
\newcommand{\mF}{\mathcal{F}}
\newcommand{\mD}{\mathcal{D}}
\newcommand{\htheta}{\hat{\theta}}
%---------------------------------------------------------------------------
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{con}[thm]{Conjecture}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{ex}[thm]{Example}
\newtheorem{qn}[thm]{Question}
\newcommand{\defn}{\bigbreak\noindent{\bf Definition. }}
\newcommand{\undefn}{\bigbreak}
\newcommand{\rmk}{\bigbreak\noindent{\bf Remark. }}
\newcommand{\unrmk}{\bigbreak}
%\newcommand{\proof}{\noindent{\bf Proof. }}
\newcommand{\unproof}{\hfill$\Box$\bigbreak}
\newcommand{\defeq}{\stackrel{\mathrm{def}}{=}}
\newcommand{\holds}[1]{\stackrel{?}{#1}}
\newcommand{\conv}{\mathrm{conv}}
\newcommand{\ie} {{\it i.e. }}
\newcommand{\eg} {{\it e.g. }}
\newcommand{\etal} {{\it et al. }}
\newcommand{\E}{\mathbf{E}}
\newcommand{\betavec}{\pmb{\beta}}
\newcommand{\qvec}{\mathbf{q}}
\newcommand{\lpf}{\mathbf{\psi}} %logpartition function
\newcommand{\family}{\mathcal F} 
\newcommand{\suff}{\pmb{\phi}} % suff stats
\newcommand{\muvec}{\pmb{\mu}}
\newcommand{\st}{\hspace{10pt}\mathrm{s.t.}\hspace{10pt}}
\newcommand{\norm}[1]{|| #1 ||}
\newcommand{\tp}{\tilde{p}}

@@

\section{Exponential Utility}

Assume the agent's belief distribution $p$ belongs to an exponential family, so it takes the form
%
\[
p(x;\theta) = \exp[\theta x - T(\theta)]
\]
%
where $\theta$ is the natural parameter and $T$ is the log-partition function. (I'm assuming that the sufficient statistic is $\phi(x) = x$ just for simplicity.) Assume also that the agent has an exponential utility for money $w$:
%
\[
U(w) = -\frac{1}{a} \exp(-aw).
\]
%
Here $a$ is the coefficient of risk aversion (higher means more risk averse, and the utility function is more concave).

%
\begin{proposition}\label{prop:exp-util}
An agent with exponential family belief with natural parameter $\htheta$, and exponential utility with coefficient $a$, makes a trade that moves the current market share vector $\theta$ to the convex combination $\frac{1}{1+a} \htheta + \frac{a}{1+a} \theta$ assuming an LMSR cost function.
\end{proposition}
%
%
\begin{proof}
Let $\delta$ be the vector of shares the agent trades. The payoff given eventual outcome $x$ is then $\delta x - C(\delta + \theta) + C(\theta)$. The utility for this payoff is as follows (recall that $\htheta$ is the agent's believed natural parameter).
%
\[ U( \delta x - C(\delta + \theta) + C(\theta) ) = -\frac{1}{a} \exp(-a \delta x + a C(\delta + \theta) - a C(\theta)).
\]
%
Taking the expected utility, we obtain
%
\begin{eqnarray*}
&   & \Exp\left[ U( \delta x - C(\delta + \theta) + C(\theta) ) \right] \\
& = & \int_{\mX} -\frac{1}{a} \exp(-a \delta x + a C(\delta + \theta) - a C(\theta)) \exp[\htheta x - T(\htheta)]\, dx \\
& = & -\frac{1}{a} \int_{\mX} \exp[(\htheta -a \delta) x + a C(\delta + \theta) - a C(\theta)) - T(\htheta)]\, dx \\
& = & -\frac{1}{a} \exp[a C(\delta + \theta) - a C(\theta)) + T(\htheta - a\delta) - T(\htheta)] \int_{\mX} \exp[(\htheta -a \delta) x - T(\htheta -a \delta)]\, dx \\
& = & -\frac{1}{a} \exp[a C(\delta + \theta) - a C(\theta)) + T(\htheta - a\delta) - T(\htheta)] \int_{\mX} p(x; \htheta -a \delta)\, dx \\
& = & -\frac{1}{a} \exp[a C(\delta + \theta) - a C(\theta)) + T(\htheta - a\delta) - T(\htheta)]\\
& = & U\left(- C(\delta + \theta) + C(\theta)) - \frac{1}{a} T(\htheta - a\delta) + \frac{1}{a} T(\htheta)\right)
\end{eqnarray*}
%
The second-last equality follows from the fact that $\int_{\mX} p(x; \htheta -a \delta)\, dx = 1$. Since utility $U$ is monotone increasing, it is maximized by maximizing its argument, which is a concave function of $\delta$ by convexity of $C$ and $T$. The optimality condition for the argument is
%
\begin{equation} \label{eq:optim}
\grad C(\delta^* + \theta) = \grad T(\htheta - a\delta^*)
\end{equation}
%
Now if the market maker is using LMSR, then $C$ is the log-partition function of the corresponding exponential family and $C = T$. Then~(\ref{eq:optim}) can be solved by equating the arguments. This leads to $\delta^* = (\htheta - \theta) / (1+a)$, which moves the share vector to $\theta + \delta^* = \frac{1}{1+a} \htheta + \frac{a}{1+a} \theta$.
\end{proof}
%

In the statement of the result I use the term ``LMSR cost function'' somewhat loosely, because we are not necessarily dealing with a market over exhaustive, mutually exclusive outcomes. What is meant is the cost function that arises by taking the dual to entropy of the maxent distribution with given mean parameter $\mu$. As we've discussed this seems like the right generalization of LMSR to arbitrary mean parameter spaces.

Note that as $a \rightarrow 0$, we approach risk neutrality and the agent moves the share vector all the way to its private estimate $\htheta$. As $a$ grows (agent grows more risk averse) the agent makes smaller and smaller trades that keep it closer to the current estimate $\theta$.

So there is a kind of congruence between exponential family beliefs and exponential utility (as the names would suggest).

>>>>

%new commands
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\mX}{\mathcal{X}}
\newcommand{\bR}{\mathbf{R}}
\newcommand{\grad}{\nabla}
\newcommand{\Exp}{\mathbf{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\mM}{\ensuremath{\mathcal{M}}}
\newcommand{\hmu}{\ensuremath{\hat{\mu}}}
\newcommand{\mP}{\ensuremath{\mathcal{P}}}
\newcommand{\mY}{\ensuremath{\mathcal{Y}}}
\newcommand{\mS}{\ensuremath{\mathcal{S}}}
\newcommand{\ds}{\displaystyle}
\newcommand{\mF}{\mathcal{F}}
\newcommand{\mD}{\mathcal{D}}
\newcommand{\htheta}{\hat{\theta}}
%---------------------------------------------------------------------------
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{con}[thm]{Conjecture}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{ex}[thm]{Example}
\newtheorem{qn}[thm]{Question}
\newcommand{\defn}{\bigbreak\noindent{\bf Definition. }}
\newcommand{\undefn}{\bigbreak}
\newcommand{\rmk}{\bigbreak\noindent{\bf Remark. }}
\newcommand{\unrmk}{\bigbreak}
%\newcommand{\proof}{\noindent{\bf Proof. }}
\newcommand{\unproof}{\hfill$\Box$\bigbreak}
\newcommand{\defeq}{\stackrel{\mathrm{def}}{=}}
\newcommand{\holds}[1]{\stackrel{?}{#1}}
\newcommand{\conv}{\mathrm{conv}}
\newcommand{\ie} {{\it i.e. }}
\newcommand{\eg} {{\it e.g. }}
\newcommand{\etal} {{\it et al. }}
\newcommand{\E}{\mathbf{E}}
\newcommand{\betavec}{\pmb{\beta}}
\newcommand{\qvec}{\mathbf{q}}
\newcommand{\lpf}{\mathbf{\psi}} %logpartition function
\newcommand{\family}{\mathcal F} 
\newcommand{\suff}{\pmb{\phi}} % suff stats
\newcommand{\muvec}{\pmb{\mu}}
\newcommand{\st}{\hspace{10pt}\mathrm{s.t.}\hspace{10pt}}
\newcommand{\norm}[1]{|| #1 ||}
\newcommand{\tp}{\tilde{p}}
\newcommand{\CE}{C\!E}

@@

\section{Risk Averse Traders with Exponential Utility}

In this section we relax that standard assumption that agents in the market are risk-neutral. We show that, with sufficient extra structure to the agents' beliefs and utilities, the market performs a clean aggregation of the agents' expectations in the form of a simple weighted averages.
Assume that the agent has an exponential utility function for money $w$:
%
\begin{equation} \label{exp-util}
U_a(w) = -\frac{1}{a} \exp(-aw).
\end{equation}
%
Here $a$ controls the level of risk aversion: the agent is more risk averse as $a$ increases, and as $a$ tends to 0 we approach linear utility (risk-neutrality). Specifically, $a$ is the Arrow-Pratt coefficient of absolute risk aversion, and exponential utilities of the form~(\ref{exp-util}) are the unique utilities that exhibit constant absolute risk aversion (CARA).

If wealth is distributed according to a probability measure $P$, then the \emph{certainty equivalent} of a random amount of wealth is defined as
$$
\CE(w) = U_a^{-1} \Exp_P\left[ U_a(w) \right].
$$
Suppose as before that the agent's belief over outcomes takes the form of a density $p$ with respect to base measure $\nu$. There is a close relationship between the log-partition function and the certainty equivalent under exponential utility.
%
\begin{lemma} \label{lem:exp-util}
The certainty equivalent of the agent's expected profit, under exponential utility, when acquiring shares $\delta$ under a market state of $\theta$ is
\begin{equation}
\log a - T_p(-a\delta) - aC(\theta+\delta) + aC(\theta),
\end{equation}
where $T_p$ is the log partition function~(\ref{log-part}) with a base measure of $p\,d\nu$. Furthermore, if the agent's belief is an exponential family with parameter $\htheta$, we have
%
\begin{equation*}
T_p(\delta) = T(\htheta + \delta) - T(\htheta),
\end{equation*}
%
where $T$ is the usual log partition function with base measure $\nu$.
\end{lemma}
%
%
\begin{proof}
Explicitly, the certainty equivalent of the profit is
%
\begin{eqnarray*}
& & \CE(\,\la\delta,\phi(x)\ra - [C(\theta+\delta) - C(\theta)]\,)\\
& = & -\log \int_{\mX} \frac{1}{a} \exp\left( \la-a\delta,\phi(x)\ra + a [C(\theta+\delta) - C(\theta)]  \right) p(x)d\nu(x)\\
& = & \log a - a [C(\theta+\delta) - C(\theta)] - \log \int_{\mX} \exp\left( \la-a\delta,\phi(x)\ra \right)  p(x)d\nu(x)\\
& = & \log a - a [C(\theta+\delta) - C(\theta)] - T_p(-a\delta).
\end{eqnarray*}
%
For the second part of the result, we have
%
\begin{eqnarray*}
T_p(\delta) & = & \log \int_{\mX} \exp(\la\delta,\phi(x)\ra)\,p(x;\htheta)\, d\nu(x)\\
& = &  \log \int_{\mX} \exp( \la\delta+\htheta,\phi(x)\ra - T(\htheta) )\, d\nu(x)\\
& = &  T(\htheta+\delta) - T(\theta) + \log \int_{\mX} \exp( \la\delta+\htheta,\phi(x)\ra - T(\htheta+\delta) )\, d\nu(x)\\
& = &  T(\htheta+\delta) - T(\theta) + \log \int_{\mX} p(x;\htheta+\delta) \, d\nu(x)\\
& = &  T(\htheta+\delta) - T(\theta),
\end{eqnarray*}
%
where the last line follows from the fact that density $p(x;\htheta+\delta)$ integrates to 1.
\end{proof}

\noindent
Recall that for the generalized LMSR, the cost function $C$ is exactly the log partition function $T$. We are therefore lead to the following clear understanding of a risk-averse agent's behavior in such a market.
%
\begin{theorem}
Suppose an agent has exponential utility with coefficient $a$ and exponential family beliefs with natural parameter $\htheta$. In the generalized LMSR market with current market state $\theta$, the agent's optimal trade $\delta$ moves the state vector to
%
\begin{equation} \label{exp-update}
\theta + \delta = \frac{1}{1+a} \htheta + \frac{a}{1+a} \theta.
\end{equation}
%
\end{theorem}
%
%
\begin{proof}
The agent's optimal trade maximizes its expected utility, or equivalently the certainty equivalent. From Lemma~\ref{lem:exp-util} and the fact that $T = C$, the agent therefore maximizes
%
$$
\log a - T(\htheta - a\delta) + T(\htheta) - aT(\theta+\delta) + aT(\theta). 
$$
This objective is strictly concave, from the strict convexity of $T$. The optimum is therefore characterized by the first-order conditions:
$$
\grad T(\htheta - a\delta) = \grad T(\theta+\delta).
$$
As the gradient map $\grad T$ is one-to-one, this is solved by equating the arguments, which leads to $\delta = (\htheta - \theta) / (1+a)$ and~(\ref{exp-update}).
% 
\end{proof}
%

Note that as, $a$ tends to 0, we approach risk neutrality and the agent moves the share vector all the way to its private estimate $\htheta$. As $a$ grows larger (agent grows more risk averse) the agent makes smaller trades to reduce it exposure, and the final state stays closer to the current state $\theta$. Update~(\ref{exp-update}) implies that under the conditions of the theorem, a market that receives a sequence of myopic traders aggregates their natural parameters in the form of an exponentially weighted moving average. The final market estimates (i.e., prices) are obtained by applying $\grad T$ to this average.
